Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-06 tex/main.tex
intro

\newcommand\cvcd{CVC4d\xspace}
\newcommand\cvcf{CVC4f\xspace}
\newcommand\cvci{CVC4i\xspace}
\newcommand\ziiib{\ziii}

@@

\begin{abstract}
SMT solvers have recently been extended with finite model finding techniques.
However, these cannot cope with universal quantifiers ranging over infinite
types. This paper introduces a translation that eliminates infinite quantifiers
in axioms specifying a large class of recursive functions, including
terminating, tail-recursive, and productive functions. An empirical evaluation
confirms that the approach works well on benchmarks from two sources. The
translation is implemented as a preprocessor in the solver CVC4.
\end{abstract}

@@

In the past few years, techniques have been developed to finding models for
quantified formulas in SMT.
Ge and de Moura \cite{GeDeM-CAV-09} introduced a complete instantiation-based
procedure for formulas in the \relax{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all
occurrences of variables occur as direct subterms of uninterpreted
functions---e.g., $\forall x%\typ{\sortint} --- the sort is not important, right?
.\;\, \con{f}( x )
\teq \con{g}( x ) + 5$.
% since all
% occurrences of $x$ occur beneath uninterpreted functions $\con{f}$ and
% $\con{g}$.
(Other syntactic criteria extend
this fragment slightly, including cases when variables occur as arguments of
arithmetic predicates.) Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013} introduced techniques for finding finite
models for quantified
formulas over uninterpreted types as well as types having a fixed finite
interpretation, such as fixed-width bit vectors and enumerated datatypes.
These techniques can
find a model for the formula $\forall x,\, y : \tau.\;\, x \teq
y \vee \neg \con{f}( x ) \teq \con{f}( y )$.
% where the uninterpreted type $\tau$ is
% interpreted as a finite set and $\con{f}$ as injective finite map.

Unfortunately, these fragments cannot accommodate the vast majority of
quantified formulas that correspond to recursive function definitions: The
essentially uninterpreted fragment does not allow the argument of a
recursive function to be used deep inside a complex term on the right-hand side,
whereas the finite model finding techniques of Reynolds et al.\ are not
applicable for functions over infinite domains such as the integers. A simple
example where both approaches fail is
$$\forall n : {\sortint}.\;\, \lite\bigl(
n \leq 0,\allowbreak\; \con{p}( n ) \teq 1,\allowbreak\; \con{p}( n ) \teq 2 * \con{p}( n - 1 ) \bigr)$$
This state of affairs is highly unsatisfactory, given the frequence of
recursive definitions in practice and the impending addition of a dedicated
command, \texttt{define-funs-rec}, to the SMT-LIB standard \cite{smtlib25}.

@@

We define an (un)satisfiability-preserving translation method for a class of
formulas containing admissible recursive function definitions
(Section~\ref{sec:encoding}) that produces formulas for which existing model
finding techniques are applicable. The main insight is that a
recursive definition $\forall x : \tau.\; \const{f}(x) \teq t$
can be translated to
$\forall a :
\fargsort{\tau}.\; \const{f}(\farg{\con{f}}(a)) \teq t[\concret(a)/x]$, where
$\fargsort{\tau}$ is an uninterpreted ``abstract'' type and $\farg{\con{f}} :
\fargsort{\tau} \to \tau$ a coercion from the abstract type to the concrete
type. The translation preserves satisfiability and, for admissible definitions,
unsatisfiability, and it makes finite model finding possible.

Our empirical evaluation on benchmarks from provides evidence that this translation improves the
effectiveness of SMT solvers \cvc and \ziii for finding counterexamples to
verification conditions (Section~\ref{sec:evaluation}).

We detail front-end support for recursive function definitions and this
translation in the SMT solver \cvc (Section~\ref{sec:front-end}).

Admissibility must be checked separately---for example, using a
syntactic criterion or a termination prover.

Usually not interested in showing logical consistency, there
exist other tools that check e.g. termination.
Thus, want SMT solver to check satisfiability the assumption that recursive
functions are consistent (where this needs to be defined precisely).

\newpage

  * quantifiers in recursive fun defs

  * intuitively, finite quantification suffices:
    * if we make sure all needed values are there
    * if well-founded (assumed!)

  * nasty example (not well-founded)
    * f x = f x + 1

  * how we do it

  * corecursive functions???
    * productive

\begin{verbatim}
this option targets users who wish to determine the satisfiability of inputs
that involving function definitions that are known to be consistent, whereas it
does not target users who wish to check the consistency of a set of function
definitions.
\end{verbatim}

@@

We gathered a set of 245 benchmarks from two sources, which we will refer to as
{\bench isa} and {\bench leon}. The first ({\bench isa}) consists of the 79
benchmarks from the Isaplanner benchmark
suite~\cite{DBLP:conf/itp/JohanssonDB10} that do not contain higher-order
functions. The benchmarks have been used recently as challenge problems for a
variety of inductive theorem provers. These benchmarks heavily involve
recursive functions, and are limited to the combined theory of uninterpreted
functions and inductive datatypes. The second ({\bench leon}) consists of 166
benchmarks from the Leon repository,\footnote{%Available at
\url{https://github.com/epfl-lara/leon}} which were constructed from
verification conditions from simple Scala programs. These benchmarks also
heavily involve recursive functions over inductive datatypes, and also cover a
wide variety of theories, including bit vectors, arrays, and linear and
nonlinear arithmetic. Each benchmark is in definitional form with respect to a
set of terminating functions.
%
% in particular they each
%are a conjoined list of function definitions followed by a (negated)
%conjecture.
A majority of these benchmarks are unsatisfiable, although a
handful of the benchmarks in the {\bench leon} set are satisfiable.

@@

\section{Front-End Support %for Recursive Function Definitions
  in CVC4}
\label{sec:front-end}

@@

The \texttt{--fmf-fun} option instructs CVC4 to assume that the
functions provided by the user are admissible.
If this condition is not met, the solver may answer ``satisfiable'' on
an unsatisfiable problem. For example, if we add the inconsistent definition
%
\begin{quote}
\begin{verbatim}
(define-fun-rec e ((n Int)) Int (+ (e n) 1))
\end{verbatim}
\end{quote}
%
to the above problem, CVC4 will still find the same ``model''
if \texttt{--fmf-fun} is specified.

@@

\begin{verbatim}
  * future work
    * more eval
      * against other non-SMT-solvers (notably Leon and Nitpick)
      * problems exercising codatatypes (which CVC4 supports \cite{xxx}) and
        corecursive functions
    * more features
      * e.g. encode (co)inductive predicates as (co)recursive functions
\end{verbatim}

>>>>

\newcommand\cvcd{{\cvc}d\xspace}
\newcommand\cvcf{{\cvc}f\xspace}
\newcommand\cvci{{\cvc}i\xspace}
\newcommand\ziiib{\ziii}

@@

\begin{abstract}
SMT solvers have recently been extended with finite model finding techniques.
However, these generally cannot cope with universal quantifiers ranging over
infinite types. This paper introduces a translation that eliminates infinite
quantifiers in axioms specifying a large class of recursive functions,
including terminating, tail-recursive, and productive functions. An empirical
evaluation confirms that the approach works well on benchmarks from two
independent sources. The translation is implemented as a preprocessor in the
solver CVC4.
\end{abstract}

@@

In the past few years, techniques have been developed to finding models for
quantified formulas in SMT.
Ge and de Moura \cite{GeDeM-CAV-09} introduced a complete instantiation-based
procedure for formulas in the \relax{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all
occurrences of variables occur as direct subterms of uninterpreted
functions---e.g., $\forall x%\typ{\sortint} --- the sort is not important, right?
.\;\, \con{f}( x )
\teq \con{g}( x ) + 5$.
% since all
% occurrences of $x$ occur beneath uninterpreted functions $\con{f}$ and
% $\con{g}$.
Other syntactic criteria extend
this fragment slightly, including cases when variables occur as arguments of
arithmetic predicates. More recently, Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013} introduced techniques for finding finite
models for quantified
formulas over uninterpreted types as well as types having a fixed finite
interpretation, such as fixed-width bit vectors and enumerated datatypes.
These techniques can
find a model for the formula $\forall x,\, y : \tau.\;\, x \teq
y \vee \allowbreak \neg \con{f}( x ) \teq \con{f}( y )$, where $\tau$ is uninterpreted.
% where the uninterpreted type $\tau$ is
% interpreted as a finite set and $\con{f}$ as injective finite map.

Unfortunately, neither of these fragments can accommodate the vast majority of
quantified formulas that correspond to recursive function definitions: The
essentially uninterpreted fragment does not allow the argument of a
recursive function to be used deep inside a complex term on the right-hand side,
whereas the finite model finding techniques of Reynolds et al.\ are not
applicable for functions over infinite domains such as the integers. A simple
example where both approaches fail is
$$\forall n : {\sortint}.\;\, \lite\bigl(
n \leq 0,\allowbreak\; \con{p}( n ) \teq 1,\allowbreak\; \con{p}( n ) \teq 2 * \con{p}( n - 1 ) \bigr)$$
This state of affairs is highly unsatisfactory, given the frequence of
recursive definitions in practice and the impending addition of a dedicated
command, \texttt{define-fun-rec}, to the SMT-LIB standard \cite{smtlib25}.

@@

We define a translation for a class of
formulas containing admissible recursive function definitions
(Section~\ref{sec:encoding}) that produces formulas for which existing model
finding techniques are applicable. The main insight is that a
recursive definition $\forall x : \tau.\; \const{f}(x) \teq t$
can be translated to
$\forall a :
\fargsort{\tau}.\; \const{f}(\farg{\con{f}}(a)) \teq t[\concret(a)/x]$, where
$\fargsort{\tau}$ is an uninterpreted ``abstract'' type and $\farg{\con{f}} :
\fargsort{\tau} \to \tau$ a coercion from the abstract type to the concrete
type. The translation preserves satisfiability and, for admissible definitions,
unsatisfiability, and it makes finite model finding possible in practice.

Our empirical evaluation on benchmarks from IsaPlanner and Leon provides
evidence that this translation improves the effectiveness of the SMT solvers
\cvc and \ziii for finding counterexamples to verification conditions
(Section~\ref{sec:evaluation}). The approach is implemented as a preprocessor
in \cvc. When invoked with the appropriate command-line option,
the solver assumes functions introduced using the \texttt{define-fun-rec} command are
admissible and translates them accordingly (Section~\ref{sec:front-end}).

%this option targets users who wish to determine the satisfiability of inputs
%that involving function definitions that are known to be consistent, whereas it
%does not target users who wish to check the consistency of a set of function
%definitions.

@@

We gathered a set of 245 benchmarks from two sources, which we will refer to as
{\bench isa} and {\bench leon}. The first ({\bench isa}) consists of the 79
benchmarks from the IsaPlanner
suite~\cite{DBLP:conf/itp/JohanssonDB10} that do not contain higher-order
functions. The benchmarks have been used recently as challenge problems for a
variety of inductive theorem provers. These benchmarks heavily involve
recursive functions, and are limited to the combined theory of uninterpreted
functions and inductive datatypes. The second ({\bench leon}) consists of 166
benchmarks from the Leon repository,\footnote{%Available at
\url{https://github.com/epfl-lara/leon}} which were constructed from
verification conditions from simple Scala programs. These benchmarks also
heavily involve recursive functions over inductive datatypes, and also cover a
wide variety of theories, including bit vectors, arrays, and linear and
nonlinear arithmetic. Each benchmark is in definitional form with respect to a
set of terminating functions.
%
% in particular they each
%are a conjoined list of function definitions followed by a (negated)
%conjecture.
A majority of these benchmarks are unsatisfiable, although a
handful of the benchmarks in the {\bench leon} set are satisfiable.

@@

\section{Front-End Support in \cvc}
\label{sec:front-end}

@@

The \texttt{--fmf-fun} option instructs \cvc to assume that the
functions provided by the user are admissible.
Admissibility must be checked separately---for example, using a
syntactic criterion or a termination prover.
If this criterion is not met, the solver may answer ``satisfiable'' on
an unsatisfiable problem. For example, if we add the inconsistent definition
%
\begin{quote}
\begin{verbatim}
(define-fun-rec e ((n Int)) Int (+ (e n) 1))
\end{verbatim}
\end{quote}
%
to the above problem, \cvc will still find the same ``model''
if \texttt{--fmf-fun} is specified.

%Usually not interested in showing logical consistency, there
%exist other tools that check e.g. termination.
%Thus, want SMT solver to check satisfiability the assumption that recursive
%functions are consistent (where this needs to be defined precisely).

@@

\begin{verbatim}
  * future work
    * more eval
      * against other non-SMT-solvers (notably Leon and Nitpick)
      * problems exercising codatatypes (which \cvc supports \cite{xxx}) and
        corecursive functions
    * more features
      * e.g. encode (co)inductive predicates as (co)recursive functions
\end{verbatim}
